---
title: "Associations of Healthcare Costs and Early Detection with Colorectal Cancer Mortality in the USA"
author: "Yulin Yuan, Ruohan Sun, Tony Lee"
format: pdf
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warning: false

#Loading library
library(car)
library(leaps)
library(tidyverse)
library(broom)
```

```{r}
#| echo: false
# Read the CSV file
data <- read.csv("data/colorectal_cancer_dataset.csv")

# Select the data in USA
usa_data <- data %>%
  filter(Country == "USA")

# Remove Patient_ID and Country
colorectal_usa <- usa_data %>%
  select(-Patient_ID, -Country)
```

## Introduction

### Study Objective

Clearly define the research questions or objectives.

### Dataset Overview

Describe:

-   Where the data comes from

-   What each variable represents

-   How the data was collected (if known)

### Motivation

Why is this question or data important/interesting?

## Analysis

### Exploratory Data Analysis (EDA)

#### Data Visualizations

```{r}

```

#### Summary Correlations

```{r}

```

#### Interpretation of Findings

Pattern, trends, suggested operations

### Model Choice and Reasoning

#### Linear Regression

explain why choose this model based on EDA and Data description

#### Data Transformations

```{r}

```

#### Model Selection

```{r}

```

### Statistical Analysis

```{r}

```

### Results Interpretation

inference results

## Conclusion

### Main Findings

Interpreting result in real-world context, careful about causality

### Limitations

-   Discuss possible sources of bias, limitations in data, model assumptions

-   Suggest improvements or next steps

### Potential Further research

Mention anything interesting you found that doesn’t fit elsewhere

## Appendix

-   Full regression output

-   Extra plots or tables not essential to the main body

-   Model selection steps
