---
title: ""
author: ""
format: pdf
editor: visual
---

```{=latex}
\begin{titlepage}
    \centering
    \vspace*{0.3\textheight}
    {\Huge\bfseries Associations of Healthcare Costs and Early Detection\\[0.5em]
    with Colorectal Cancer Mortality in Canada \par}
    \vspace{2cm}
    {\Large Yulin Yuan, Ruohan Sun, Tony Lee\par}
    \vspace{1cm}
    {\large April 4, 2025\par}
    \vfill
\end{titlepage}
```

```{r}
#| echo: false
#| message: false
#| warning: false

#Loading library
library(car)
library(leaps)
library(tidyverse)
library(broom)
library(grid)
library(gridExtra)
library(ggplotify)
```

```{r}
#| echo: false
# Read the CSV file
data <- read.csv("data/colorectal_cancer_dataset.csv")

# Select the data in USA
ca_data <- data %>%
  filter(Country == "Canada")

# Remove Patient_ID, Country, Country-level features and other response features
colorectal_CA <- ca_data %>%
  select(
    -Patient_ID,                     # Not relevant
    -Country,                        # Focus on Canada only
    -Incidence_Rate_per_100K,        # Country-level
    -Mortality_Rate_per_100K,        # Country-level
    -Economic_Classification,        # Country-level
    -Healthcare_Access,              # Country-level
    -Survival_Prediction,            # Same effect as response variable: Mortality
    -Survival_5_years                # Same effect as response variable: Mortality
  )
```

\newpage

## Introduction

### Study Objective

Clearly define the research questions or objectives.

### Dataset Overview

The dataset we decide to conduct our analysis of Mortality status (Alive/Dead) was sourced from Kaggle, which the dataset name is “Colorectal Cancer Global Dataset & Predictions”. For the objective of this project, we only considered the data from Canada, which includes patient demographics, lifestyle risks, medical history, cancer stage, treatment types, survival chances, and healthcare costs. Key variables of we are going to investigate as follows:

#### Response Variable

\begin{longtable}{|p{4cm}|p{9cm}|p{4cm}|}
\hline
\textbf{Variable} & \textbf{Description} & \textbf{Type} \\
\hline
Mortality Status & Status of patients' mortality (Yes/No) & Categorical, nominal \\
\hline
\end{longtable}

#### Explanatory Variables

\begin{longtable}{|p{4cm}|p{9cm}|p{4cm}|}
\hline
\textbf{Variable} & \textbf{Description} & \textbf{Type} \\
\hline
Alcohol Consumption & Status of patient's alcohol consumption (Yes/No) & Categorical, nominal \\
\hline
Age & Patient's age in years & Numerical, continuous \\
\hline
Cancer Stage & Diagnosis stage of cancer (Localized, Regional, Metastatic) & Categorical, ordinal \\
\hline
Diabetes & Status of patient's diabetes (Yes/No) & Categorical, nominal \\
\hline
Diet Risk & Level of dietary risk (Low, Moderate, High) & Categorical, ordinal \\
\hline
Early Detection & Detection of colorectal cancer at an early stage (Yes/No) & Categorical, nominal \\
\hline
Family History & Presence of family history of colorectal cancer (Yes/No) & Categorical, nominal \\
\hline
Genetic Mutation & Presence of genetic mutations for colorectal cancer (Yes/No) & Categorical, nominal \\
\hline
Gender & Gender of the patient (Male/Female) & Categorical, nominal \\
\hline
Healthcare Cost & Estimated healthcare expenditure per patient (1,000 units in \$) & Numerical, continuous \\
\hline
Inflammatory Bowel Disease & Status of inflammatory bowel disease (Yes/No) & Categorical, nominal \\
\hline
Insurance Status & Health insurance coverage (Insured/Uninsured) & Categorical, nominal \\
\hline
Obesity BMI & BMI classification (Normal, Overweight, Obese) & Categorical, ordinal \\
\hline
Physical Activity & Level of physical activity (Low, Moderate, High) & Categorical, ordinal \\
\hline
Screening History & Cancer screening history (Regular, Irregular, Never) & Categorical, ordinal \\
\hline
Smoking History & Patient's smoking history (Yes/No) & Categorical, nominal \\
\hline
Treatment Type & Type of treatment received (Surgery, Chemotherapy, Radiotherapy, Combination) & Categorical, ordinal \\
\hline
Tumor Size & Colorectal tumor size in millimeters (mm) & Numerical, continuous \\
\hline
Urban or Rural & Patient's area of residence (Urban/Rural) & Categorical, nominal \\
\hline
\end{longtable}

### Motivation

-   Why is this question or data important/interesting?

-   Need to explain why we remove country-level variable and Patient_ID, reasonable is fine.

## Analysis

### Exploratory Data Analysis (EDA)

```{r}
#| echo: false
#| message: false
#| warning: false

# Set response variable 
resp <- "Mortality"
# Set response variable into factor type
colorectal_CA[[resp]] <- as.factor(colorectal_CA[[resp]])

# Set the unit of Healthcare_Costs in $1,000
colorectal_CA[["Healthcare_Costs"]] <- colorectal_CA[["Healthcare_Costs"]]/1000

# Initialize containers
continuous <- c()
categorical <- c()

# Loop through all variables
for (colname in names(colorectal_CA)) {
  if (colname == resp) next  # Skip the response variable itself
  
  column <- colorectal_CA[[colname]]
  
  if (is.numeric(column)) {
    continuous <- c(continuous, colname)
  } else if (is.character(column) || is.factor(column)) {
    categorical <- c(categorical, colname)
  }
}

# Convert all categorical variables to factors
for (colname in categorical) {
  colorectal_CA[[colname]] <- as.factor(colorectal_CA[[colname]])
}
```

#### Balance of Response Variable

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 3

# Bar plot of Mortality in original dataset
plot_not_balance <- ggplot(colorectal_CA, aes(x = Mortality, fill = Mortality)) +
  geom_bar(alpha = 0.6) +
  scale_fill_manual(values = c("blue", "red"), labels = c("Alive", "Dead")) +
  labs(
    title = "Class Distribution of Mortality (Imbalance)",
    x = "Mortality Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "none"
  )

# Count how many in minority class
min_class_size <- min(table(colorectal_CA$Mortality))

# Sample from each class equally
set.seed(123)
balanced_data <- colorectal_CA %>%
  group_by(Mortality) %>%
  sample_n(min_class_size) %>%
  ungroup()

# Bar plot of Mortality in balanced dataset
plot_balance <- ggplot(balanced_data, aes(x = Mortality, fill = Mortality)) +
  geom_bar(alpha = 0.6) +
  scale_fill_manual(values = c("blue", "red"), labels = c("Alive", "Dead")) +
  labs(
    title = "Class Distribution of Mortality (Balance)",
    x = "Mortality Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "none"
  )

my_list <- list(plot_not_balance, plot_balance)

combined_plot <- do.call(arrangeGrob, c(my_list, ncol = 2))

caption_grob <- textGrob("Figure 1: Class Distribution of the Mortality Variable Before and After Balancing",
                         gp = gpar(fontface = "bold.italic", fontsize = 15), 
                         hjust = 0.5,
                         vjust = 0.3)

final_plot <- arrangeGrob(combined_plot, bottom = caption_grob)

as.ggplot(final_plot)
```

#### Continuous Variables

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 15
#| fig-height: 5

# Create Histogram plots for Continuous Covariates of Binary Response
numeric_cols <- c("Age", "Tumor Size (mm)", "Healthcare Costs ($K)")
my_list <- list()

for (i in seq_along(continuous)) {
  feat <- continuous[i]
  feat_plot <- ggplot(balanced_data, aes_string(x = feat, fill = resp)) + 
    geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
    labs(
      #title = paste("Histogram of", numeric_cols[i]), 
      x = numeric_cols[i], 
      y = "Frequency"
    ) +
    scale_fill_manual(values = c("blue", "red"), labels = c("Alive", "Dead")) +
    guides(fill = guide_legend(title = "Mortality")) +
    theme_minimal(base_size = 16) +
    theme(
      #plot.title = element_text(hjust = 0.5, size = 18),
      axis.title = element_text(size = 16),
      axis.text = element_text(size = 14),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 14)
    )
  my_list[[i]] <- feat_plot
}

combined_plot <- do.call(arrangeGrob, c(my_list, ncol = 3))

caption_grob <- textGrob("Figure 2: Distribution of continuous variables by mortality status",
                         gp = gpar(fontface = "bold.italic", fontsize = 23), 
                         hjust = 0.5,
                         vjust = 0.3)

final_plot <- arrangeGrob(combined_plot, bottom = caption_grob)

as.ggplot(final_plot)
```

#### Categorical Variables

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 40
#| fig-height: 30

# Create Proportion Bar plots for Categorical Covariates of Binary Response
categorical_cols <- c("Gender", "Cancer Stage", "Family History", "Smoking History", "Alcohol Consumption", "Obesity BMI", "Diet Risk", "Physical Activity", "Diabetes", "Inflammatory Bowel Disease", "Genetic Mutation", " Screening History", "Early Detection", "Treatment Type", "Urban or Rural", "Insurance Status")
my_list <- list()
for (i in seq_along(categorical)) {
  feat <- categorical[i]
  feat_plot <- ggplot(balanced_data, aes_string(x = feat, fill = resp)) +
    geom_bar(position = "fill", , alpha = 0.6) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      #title = paste("Mortality Proportion by", categorical_cols[i]),
      x = categorical_cols[i],
      y = "Proportion"
    ) +
    scale_fill_manual(values = c("blue", "red"), labels = c("Alive", "Dead")) +
    guides(fill = guide_legend(title = "Mortality")) +
    theme_minimal(base_size = 32) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36),
      axis.title = element_text(size = 32),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 28),
      axis.text.y = element_text(size = 28),
      legend.title = element_text(size = 30),
      legend.text = element_text(size = 28)
    )
  
  my_list[[i]] <- feat_plot
}

combined_plot <- do.call(arrangeGrob, c(my_list, ncol = 4))

caption_grob <- textGrob("Figure 3: Proportion Bar plots for Categorical variables by mortality status",
                         gp = gpar(fontface = "bold.italic", fontsize = 62), 
                         hjust = 0.5,
                         vjust = 0.3)

final_plot <- arrangeGrob(combined_plot, bottom = caption_grob)

as.ggplot(final_plot)
```

#### Interpretation of Findings

Pattern, trends, suggested operations

### Model Choice and Reasoning

#### Logistic Regression

explain why choose this model based on EDA and Data description

#### Assumption Check

```{r}
#| echo: false
#| message: false
#| warning: false

# Fit the full logistic regression model with all predictors
full_model <- glm(Mortality ~ ., data = balanced_data, family = binomial)
```

1.  **Binary Response**

    Based on *Figure 1* , the response variable is binary

2.  **Independence**

    ```{r}
    #| echo: false
    #| message: false
    #| warning: false

    # Check for duplicate rows
    dup_check <- any(duplicated(balanced_data)) 
    ```

    No duplicate rows, independence hold

3.  **Variance Structure**

    ```{r}
    #| echo: false
    #| message: false
    #| warning: false
    #| fig-width: 7
    #| fig-height: 3

    # Compute predicted probabilities
    fitted_probs <- fitted(full_model)

    # Compute binomial variances
    binomial_var <- fitted_probs * (1 - fitted_probs)

    # Create the base plot
    variance_plot <- ggplot(data.frame(fitted_probs, binomial_var), aes(x = fitted_probs, y = binomial_var)) +
      geom_point(color = "darkgreen", alpha = 0.6) +
      geom_vline(xintercept = 0.5, linetype = "dashed", color = "red") +
      labs(
        x = expression("Predicted Probability (" * hat(pi) * ")"),
        y = expression("Binomial Variance " * hat(pi) * " (1 - " * hat(pi) * ")")
      ) +
     scale_x_continuous(
      breaks = seq(0.45, 0.55, by = 0.01), limits = c(0.44, 0.56)) +
      theme_minimal(base_size = 13)

    caption_grob <- textGrob("Figure 4: Variance peaks at predicted probability = 0.5.",
                             gp = gpar(fontface = "bold.italic", fontsize = 11), 
                             hjust = 0.5, vjust = 0.3)

    final_plot <- arrangeGrob(variance_plot, bottom = caption_grob)

    as.ggplot(final_plot)
    ```

4.  **Linearity**

    ```{r}
    #| echo: false
    #| message: false
    #| warning: false
    #| fig-height: 5
    #| fig-width: 10

    # Get the logit (link) values
    logit_vals <- predict(full_model, type = "link")

    # Create list to store plots
    logit_plot_list <- list()

    # Loop over each continuous variable
    for (i in seq_along(continuous)) {
      feat <- continuous[i]
      label <- numeric_cols[i]

      linearity_plot <- ggplot(balanced_data, aes_string(x = feat, y = logit_vals)) +
        geom_point(alpha = 0.4, color = "darkblue") +
        geom_smooth(method = "loess", color = "red", se = FALSE, linetype = "solid") +
        labs(
          #title = paste("Logit vs", label),
          x = label,
          y = expression("Logit")
        ) +
        theme_minimal(base_size = 16) +
        theme(
          #plot.title = element_text(hjust = 0.5, size = 18),
          axis.title = element_text(size = 16),
          axis.text = element_text(size = 14)
        )

      logit_plot_list[[i]] <- linearity_plot
    }

    combined_plot <- do.call(arrangeGrob, c(logit_plot_list, ncol = 3))

    caption_grob <- textGrob(
      "Figure 5: Linearity check — logit of mortality plotted against continuous variables",
      gp = gpar(fontface = "bold.italic", fontsize = 17), 
      hjust = 0.5,
      vjust = 0.3
    )

    final_plot <- arrangeGrob(combined_plot, bottom = caption_grob)

    as.ggplot(final_plot)
    ```

#### Feature Selection

```{r}
#| echo: false
#| message: false
#| warning: false

# Start with full model
selected_vars <- setdiff(names(balanced_data), resp)
current_formula <- as.formula(paste(resp, "~", paste(selected_vars, collapse = " + ")))
current_model <- glm(current_formula, data = balanced_data, family = binomial)
current_aic <- AIC(current_model)

# Store best models
best_models <- data.frame(
  Num_Covariates = length(selected_vars),
  AIC = current_aic,
  Variables = paste(selected_vars, collapse = " + "),
  stringsAsFactors = FALSE
  )

step <- 19

while (step > 1) {
  aic_list <- c()

  # Try removing each variable one at a time
  for (var in selected_vars) {
    test_vars <- setdiff(selected_vars, var)
    test_formula <- as.formula(paste(resp, "~", paste(test_vars, collapse = " + ")))
    test_model <- glm(test_formula, data = balanced_data, family = binomial)
    aic_list[var] <- AIC(test_model)
  }

  best_var_to_remove <- names(which.min(aic_list))
  best_aic <- min(aic_list)

  selected_vars <- setdiff(selected_vars, best_var_to_remove)
  current_aic <- best_aic
  current_formula <- as.formula(paste(resp, "~", paste(selected_vars, collapse = " + ")))
  current_model <- glm(current_formula, data = balanced_data, family = binomial)

  # Log best model at this level
  best_models <- rbind(best_models, data.frame(
    Num_Covariates = length(selected_vars),
    AIC = best_aic,
    Variables = paste(selected_vars, collapse = " , ")
    ))
  step <- step - 1
}

# Add the null model
best_models <- rbind(best_models, data.frame(
  Num_Covariates = 0,
  AIC = AIC(glm(Mortality ~ 1, data = balanced_data, family = binomial)),
  Variables = paste(selected_vars, collapse = " , ")
  ))

aic_plot <- ggplot(best_models, aes(x = Num_Covariates, y = AIC)) +
  geom_line(color = "black") +
  geom_point(color = "red") +
  labs(
    x = "Number of Covariates",
    y = "AIC"
  ) +
  theme_minimal() + 
  scale_x_continuous(breaks = best_models$Num_Covariates)

# Caption as a grob (bold + italic)
caption_grob <- textGrob("Figure 6: AIC vs Number of Covariates – demonstrating backward feature selection.",
                         gp = gpar(fontface = "bold.italic", fontsize = 9),
                         hjust = 0.5, vjust = 0.3)

# Combine plot and caption
final_plot <- arrangeGrob(aic_plot, bottom = caption_grob)

# Convert to ggplot object
as.ggplot(final_plot)
```

```{r}
#| echo: false
#| message: false
#| warning: false

# Fit full model
full_model <- glm(Mortality ~., data = balanced_data, family = binomial)

# Perform backward selection
backward_model <- step(full_model, direction = "backward", trace = 0)

# View summary of selected model
summay_backward <- summary(backward_model)
summary_full <- summary(full_model)
```

```{=latex}
\begin{table}[ht]
\centering
\begin{tabular}{lcc}
\hline
\textbf{Model} & \textbf{Residual Deviance} & \textbf{AIC} \\
\hline
Full Model              & 9710.7 (df = 6997) & 9764.7 \\
Backward-Selected Model & 9725.6 (df = 7019) & 9735.6 \\
\hline
\end{tabular}
\caption{Residual Deviance and AIC between the backward-selected and full models}
\label{tab:model_comparison}
\end{table}
```

...

### Statistical Analysis

```{r}
#| echo: false
#| message: false
#| warning: false

# Add losing covarite of interest back to backward selected model
reduced_additive_model <- glm(
  formula = Mortality ~ Cancer_Stage + Family_History + Healthcare_Costs + Early_Detection,
  family = binomial, data = balanced_data
  )

# fit interaction term with two covariates of interest
reduced_interaction_model <- glm(
  formula = Mortality ~ Cancer_Stage + Family_History + Healthcare_Costs * Early_Detection,
  family = binomial, data = balanced_data
  )

# Use ANOVA for goodness-of-fit test
lrt_result <- anova(reduced_additive_model, reduced_interaction_model, test = "LRT")

# Summary result for additive model
result_additive <- summary(reduced_additive_model)
```

```{=latex}
\begin{table}[htbp]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Test Statistic} & \textbf{Degrees of Freedom} & \textbf{p-value} \\
\hline
$\chi^2 = 0.549$ & 1 & 0.4586 \\
\hline
\end{tabular}
\caption{Wilks' likelihood ratio test comparison for Additive and Interaction}
\label{tab:wilks-lrt}
\end{table}
```

The result suggests the interaction term does not significantly improve model fit.

...

```{=latex}
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrl}
\hline
\textbf{Predictor} & \textbf{Estimate} & \textbf{Std. Error} & \textbf{z value} & \textbf{Pr($>|z|$)} & \textbf{Significance} \\
\hline
(Intercept)              & 0.2256713  & 0.0804475  & 2.805  & 0.00503  & ** \\
Cancer\_StageMetastatic & -0.0707315 & 0.0648033  & -1.091 & 0.27506  &    \\
Cancer\_StageRegional   & -0.1151017 & 0.0537461  & -2.142 & 0.03223  & *  \\
Family\_HistoryYes      & -0.0817314 & 0.0520083  & -1.572 & 0.11607  &    \\
Healthcare\_Costs       & -0.0018622 & 0.0008735  & -2.132 & 0.03301  & *  \\
Early\_DetectionYes     & -0.0081611 & 0.0487565  & -0.167 & 0.86707  &    \\
\hline
\end{tabular}
\caption{Regression output with significance levels: * p < 0.05, ** p < 0.01}
\label{tab:regression_output}
\end{table}
```

...

### Results Interpretation

inference results

## Conclusion

### Main Findings

Interpreting result in real-world context, careful about causality

### Limitations

-   Discuss possible sources of bias, limitations in data, model assumptions

-   Suggest improvements or next steps

### Potential Further research

Mention anything interesting you found that doesn’t fit elsewhere

## Appendix

-   Full regression output

-   Extra plots or tables not essential to the main body

-   Model selection steps
